---
title: "Creating custom CGFs (from vectorized functions)"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Creating custom CGFs (from vectorized functions)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = TRUE
)
library(saddlepoint)
```

This vignette shows how to build a new `CGF` object when you can write down:

- the CGF $K(t;\theta)$ and its derivatives (at least up to order 4), and
- (optionally) the domain constraint(s) for $t$.

The recommended constructor for most univariate or elementwise models is:

- `createCGF_fromVectorisedFunctions()`

where you provide vectorized derivative functions that operate elementwise on `tvec`.

## What you must provide

For `createCGF_fromVectorisedFunctions()`, you provide 5 vectorized functions:

- `K_vectorized_func(tvec, param)` returning a numeric vector of length `length(tvec)`
- `K1_vectorized_func(tvec, param)` returning a numeric vector
- `K2_vectorized_func(tvec, param)` returning a numeric vector (diagonal of the Hessian)
- `K3_vectorized_func(tvec, param)` returning a numeric vector
- `K4_vectorized_func(tvec, param)` returning a numeric vector

The constructor then builds:

- `K(tvec,param)` as `sum(K_vectorized_func(...))`,
- `K2(tvec,param)` as a diagonal matrix with entries `K2_vectorized_func(...)`,
- and default higher-order operators needed for the correction term.

Often very useful:

- `ineq_constraint_func(tvec,param)` returning a numeric vector of constraints $g(t,\theta)$ with feasibility defined by $g \le 0$,
- `analytic_tvec_hat_func(x,param)` if you can solve $K_1(\hat t;\theta)=x$ in closed form,
- specialized `K2_solve` / `logdetK2` for performance.

## Example: a Normal CGF (SPA is exact)

Let $X \sim \mathrm{N}(\mu,\sigma^2)$. The CGF is
$$
K(t;\mu,\sigma^2) = \mu t + \frac{1}{2}\sigma^2 t^2,
$$
and the derivatives are
$$
K_1(t) = \mu + \sigma^2 t,\quad
K_2(t) = \sigma^2,\quad
K_3(t)=0,\quad
K_4(t)=0.
$$

Below we implement a Normal CGF parameterized by $(\mu, \log\sigma^2)$ to ensure positivity of $\sigma^2$.

```{r}

NormalLogSigma2CGF <- createCGF_fromVectorisedFunctions(
  K_vectorized_func = function(tvec, param) {
    mu <- param[1]
    s2 <- exp(param[2])
    mu * tvec + 0.5 * s2 * tvec * tvec
  },
  K1_vectorized_func = function(tvec, param) {
    mu <- param[1]
    s2 <- exp(param[2])
    mu + s2 * tvec
  },
  K2_vectorized_func = function(tvec, param) {
    s2 <- exp(param[2])
    rep(s2, length(tvec))
  },
  K3_vectorized_func = function(tvec, param) rep(0, length(tvec)) * param[1],
  K4_vectorized_func = function(tvec, param) rep(0, length(tvec)) * param[1],

  # Optional: analytic saddlepoint t-hat for K1(t)=x
  analytic_tvec_hat_func = function(x, param) {
    mu <- param[1]
    s2 <- exp(param[2])
    (x - mu) / s2
  },

  op_name = "NormalLogSigma2CGF"
)

# Simulate iid data
set.seed(1)
B <- 50
mu_true <- 1.5
s2_true <- 2.0
y <- rnorm(B, mean = mu_true, sd = sqrt(s2_true))

# iid CGF for (Y_1,...,Y_B)
cgB <- iidReplicatesCGF(NormalLogSigma2CGF, iidReps = B, block_size = 1)

# Saddlepoint MLE (should match exact normal MLE very closely; for Normal SPA is exact)
fit <- find.saddlepoint.MLE(
  observed.data  = y,
  cgf            = cgB,
  starting.theta = c(0.01, log(1.3)),
  method         = "two_step"
)

# Exact MLE under Normal(mu, sigma^2):
mu_hat_exact <- mean(y)
s2_hat_exact <- mean((y - mu_hat_exact)^2)

cbind(
  spa_mle   = fit$MLEs.theta,
  exact_mle = c(mu_hat_exact, log(s2_hat_exact)),
  true      = c(mu_true, log(s2_true))
)
```

## Adding a domain constraint (when needed)

Some distributions have a restricted CGF domain in $t$.
If $K(t;\theta)$ is only defined when $g(t,\theta)\le 0$, implement:

```r

ineq_constraint_func = function(tvec, param) {
  # return a numeric vector; feasible if all entries <= 0
}
```

Example: if a univariate CGF is defined for $t < b(\theta)$, you can return `tvec - b(theta)`.

