---
title: "Additional information on EsscherTiltCGF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional information on EsscherTiltCGF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment  = "#>",
message  = FALSE,
warning  = TRUE
)

library(saddlepoint)
```


`EsscherTiltCGF()` defines exponential tilting as a CGF-level transformation:
$
K_h(t) = K(t+h) - K(h),
$
which corresponds to a change-of-measure
$
\mathrm{d}\mathbb P_h(x) \;\propto\; e^{h^\top x}\,\mathrm{d}\mathbb P(x).
$
This is not a deterministic transformation of draws from the base distribution.
In general, even though the CGF determines the law, there is no universal, closed-form exact sampler that can be constructed from CGF evaluations alone.


For many families (Normal, Poisson, Gamma, ...) an exact tilted law exists in closed form.
But for a general composite CGF object, there is no universal exact sampler.

We do not provide a generic exact `rsim()` implementation for `EsscherTiltCGF(..., cgf = ..., )`.
However, if you can simulate from the base model (`cgf$rsim()` exists), you can still simulate from the tilted model using standard Monte Carlo ideas:

- (A) Importance sampling + resampling (IS-R)  
  Draw many proposals \(X^{(m)} \sim \mathbb P\), weight them by \(w_m \propto e^{h^\top X^{(m)}}\),
  and resample.
- (B) Independence Metropolis–Hastings (IMH)  
  Use proposal \(q=\mathbb P\) (simulate from the base), and accept with probability
  \[
  \alpha(x\to x') = \min\{1,\exp(h^\top(x' - x))\}.
  \]
  
#### Helper functions

```{r}
# Convert a draw to a d x B matrix (columns = draws)
.as_draw_matrix <- function(X) {
  if (is.null(dim(X))) {
    matrix(as.numeric(X), nrow = 1L)
  } else {
    X
  }
}

# Broadcast a scalar tilt to length d if needed (otherwise require exact match)
.broadcast_tilt <- function(tilt, d) {
  if (!is.numeric(tilt) || length(tilt) < 1L) stop("'tilt' must be a numeric vector.")
  if (any(!is.finite(tilt))) stop("'tilt' must be finite.")
  if (length(tilt) == 1L && d > 1L) return(rep(tilt, d))
  if (length(tilt) != d) stop("tilt length mismatch: expected length ", d, ", got ", length(tilt), ".")
  tilt
}

# Importance sampling + resampling
esscher_rsim_is_resample <- function(base_cgf,
                                     tilt,
                                     iidReps,
                                     parameter_vector,
                                     proposal_iidReps = max(2000L, 20L * iidReps),
                                     drop = TRUE,
                                     return_details = FALSE,
                                     ...) {
  if (!inherits(base_cgf, "CGF")) stop("'base_cgf' must be a CGF.")
  if (!isTRUE(base_cgf$has_simulate())) stop("Base CGF does not support simulation (has_simulate() is FALSE).")
  if (!is.numeric(iidReps) || length(iidReps) != 1L || iidReps < 1L) stop("'iidReps' must be a positive integer-ish.")
  iidReps <- as.integer(iidReps)
  proposal_iidReps <- as.integer(proposal_iidReps)

  Xprop <- .as_draw_matrix(base_cgf$rsim(iidReps = proposal_iidReps,
                                        parameter_vector = parameter_vector,
                                        drop = FALSE, ...))
  d <- nrow(Xprop)
  h <- .broadcast_tilt(tilt, d)

  # log-weights: lw_m = h^T x_m
  lw <- as.numeric(crossprod(h, Xprop))
  mx <- max(lw)
  if (!is.finite(mx)) stop("Non-finite tilt weights; try a smaller |tilt| or check your model.")

  w <- exp(lw - mx)
  sw <- sum(w)
  if (!is.finite(sw) || sw <= 0) stop("Degenerate importance weights; try more proposals or a smaller |tilt|.")
  w <- w / sw

  # Effective sample size (ESS) for diagnostics
  ess <- 1 / sum(w^2)

  idx <- sample.int(ncol(Xprop), size = iidReps, replace = TRUE, prob = w)
  out <- Xprop[, idx, drop = FALSE]

  if (drop && d == 1L) out <- as.numeric(out)

  if (isTRUE(return_details)) {
    return(list(draws = out, ess = ess, weights = w, logw = lw))
  }
  out
}

# Independence Metropolis–Hastings with proposal q = base distribution
esscher_rsim_imh <- function(base_cgf,
                             tilt,
                             iidReps,
                             parameter_vector,
                             burnin = 1000L,
                             thin = 1L,
                             init = NULL,
                             drop = TRUE,
                             return_details = FALSE,
                             ...) {
  if (!inherits(base_cgf, "CGF")) stop("'base_cgf' must be a CGF.")
  if (!isTRUE(base_cgf$has_simulate())) stop("Base CGF does not support simulation (has_simulate() is FALSE).")
  if (!is.numeric(iidReps) || length(iidReps) != 1L || iidReps < 1L) stop("'iidReps' must be a positive integer-ish.")
  iidReps <- as.integer(iidReps)
  burnin <- as.integer(burnin)
  thin <- as.integer(thin)
  if (burnin < 0L) stop("'burnin' must be >= 0.")
  if (thin < 1L) stop("'thin' must be >= 1.")

  # Infer dimension d from one base draw
  x1 <- .as_draw_matrix(base_cgf$rsim(iidReps = 1L, parameter_vector = parameter_vector, drop = FALSE, ...))
  d <- nrow(x1)
  h <- .broadcast_tilt(tilt, d)

  # Initialize chain
  if (is.null(init)) {
    x_curr <- as.numeric(x1)
  } else {
    init <- as.numeric(init)
    if (length(init) != d) stop("'init' must have length d = ", d, ".")
    x_curr <- init
  }

  n_iter <- burnin + thin * iidReps
  out <- matrix(NA_real_, nrow = d, ncol = iidReps)

  n_acc <- 0L
  k <- 0L

  for (iter in seq_len(n_iter)) {
    x_prop <- as.numeric(.as_draw_matrix(
      base_cgf$rsim(iidReps = 1L, parameter_vector = parameter_vector, drop = FALSE, ...)
    ))

    # log acceptance ratio: h^T(x_prop - x_curr)
    log_alpha <- sum(h * (x_prop - x_curr))

    # accept if log(U) < min(0, log_alpha)
    if (log(runif(1)) < min(0, log_alpha)) {
      x_curr <- x_prop
      n_acc <- n_acc + 1L
    }

    if (iter > burnin && ((iter - burnin) %% thin == 0L)) {
      k <- k + 1L
      out[, k] <- x_curr
    }
  }

  acc_rate <- n_acc / n_iter
  if (drop && d == 1L) out <- as.numeric(out)

  if (isTRUE(return_details)) {
    return(list(draws = out, accept_rate = acc_rate))
  }
  out
}
```


#### Simple example: saddlepoint MLE

We use a *simple* base model (Poisson) with a known closed-form tilt.
The exact tilted law is:
if \(X \sim \mathrm{Pois}(\lambda)\), then under tilt \(\tau\),
\(X^{(\tau)} \sim \mathrm{Pois}(\lambda e^\tau)\).

Even though we could sample exactly here, the same IS-R / IMH code applies
to more complex composite CGFs whenever `base_cgf$rsim()` exists.

```{r}
set.seed(1)

# Base model: Poisson with theta = lambda
theta_true <- c(lambda = 10)
tau <- 0.4

base <- PoissonModelCGF(lambda = adaptor(indices = 1))
tilted_cgf <- EsscherTiltCGF(base, tilt = tau)

# mean under the tilt is K'_base(tau) = lambda * exp(tau)
mean_theory <- as.numeric(base$K1(tau, theta_true))

# --- (A) 
sim_is <- esscher_rsim_is_resample(
  base_cgf = base,
  tilt = tau,
  iidReps = 2000,
  parameter_vector = theta_true,
  proposal_iidReps = 20000,
  return_details = TRUE
)

mean(sim_is$draws)
mean_theory
sim_is$ess  # diagnostic; higher is better (<= proposal_iidReps)

# --- (B) 
sim_mh <- esscher_rsim_imh(
  base_cgf = base,
  tilt = tau,
  iidReps = 2000,
  parameter_vector = theta_true,
  burnin = 2000,
  thin = 2,
  return_details = TRUE
)

mean(sim_mh$draws)
mean_theory
sim_mh$accept_rate
```

Now we fit \(\lambda\) by saddlepoint MLE under the tilted model, using the simulated tilted sample as data.

```{r}
# Using the IS-R draws as "observed data" under the tilted model
y_obs <- sim_is$draws
fit <- find.saddlepoint.MLE(
  observed.data  = y_obs,
  cgf            = tilted_cgf,
  starting.theta = c(lambda = 1),
  lb.theta       = c(lambda = 1e-8),
  ub.theta       = c(lambda = 100),
  method         = "two_step",
  std.error      = FALSE
)
fit$MLEs.theta
theta_true
```

> **Notes**
>
> - IS-R can degenerates if the tilt is large (check ESS).
> - IMH is often more robust for large tilts, but you should still check mixing.
> - If your base family has a closed-form tilted law (Normal / Poisson / Gamma / ...), prefer that exact sampler. The generic IS-R / IMH methods are fallbacks.








