---
title: "Random simulation with CGF objects: rsim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random simulation with CGF objects: rsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = TRUE
)

library(saddlepoint)
```

This vignette documents how `rsim(...)` works across CGF objects and wrappers in this codebase.
We will keep extending it as new CGFs gain simulation support.

> **Conventions used here**
>
> - `iidReps` always means “how many independent draws to simulate”.
> - For a `d`-dimensional random vector, simulation returns a `d x (iidReps)` matrix where each **column** is one draw.
> - If `d == 1` and `drop = TRUE` (default), simulation returns a numeric vector of length `iidReps`.


## What `block_size` is, and when you need to think about it

A lot of CGFs in this `saddlepoint` package are *vectorized*. That means a long `tvec` could mean either:

- **one** high-dimensional random vector, or
- **B** independent “blocks” of a smaller random vector (i.i.d. replication).

The wrapper `iidReplicatesCGF()` disambiguates those interpretations. It uses:

- `block_size` = dimension of one observation (one block), and
- `iidReps` (inside the wrapper) = how many blocks to expect (or `"any"` to infer it from `length(tvec)`).

### Does `block_size` matter for `rsim()`?

Usually, **no**. `rsim(iidReps = B, ...)` is unambiguous: you asked for `B` draws, and simulation can determine the dimension `d` from the model parameters.

You mainly need `block_size` when *constructing* a CGF (e.g., RSS models) so that evaluation methods (`K`, `K1`, `K2`, etc.) cannot silently reinterpret a long `tvec` as “a different statistical model”.

Wrappers like `linearlyMappedCGF()` can usually infer `block_size` internally (e.g., `nrow(A)`), so you rarely worry about it there.

## Quick checklist for `rsim()` support

A CGF supports simulation if:

```{r}
# cgf$has_simulate()
```

returns `TRUE`.

When it does, the canonical call is:

```r
Y <- cgf$rsim(iidReps = 10, parameter_vector = theta)
```

## Example 1a: PoissonCGF (scalar and vector)

```{r}
set.seed(1)

# Scalar Poisson (d = 1)
lambda <- 2
x <- PoissonCGF$rsim(iidReps = 10, parameter_vector = lambda)
x

# Vector Poisson (d = 2) interpreted as independent components
lambda2 <- c(2, 3)
X <- PoissonCGF$rsim(iidReps = 5, parameter_vector = lambda2, drop = FALSE)
X
```

## Example 1b: NormalCGF (vectorized independent normals)

```{r}
set.seed(1)

# Two independent normals packed into one CGF parameter vector:
# parameter_vector = c(mu1, mu2, sigma1, sigma2)
param <- c(0, 5, 1, 2)

X <- NormalCGF$rsim(iidReps = 5, parameter_vector = param, drop = FALSE)
X
dim(X)  # 2 x 5
rowMeans(X)  # ~ c(0, 5)
```

## Example 1c: BinomialCGF (vectorized)

```{r}
set.seed(1)

# Two independent binomials:
# parameter_vector = c(n1, n2, p1, p2)
param <- c(10, 20, 0.2, 0.6)

X <- BinomialCGF$rsim(iidReps = 6, parameter_vector = param, drop = FALSE)
X
rowMeans(X)              # ~ c(2, 12)
rowMeans(X) / c(10, 20)  # ~ c(0.2, 0.6)
```

## Example 2: linearlyMappedCGF simulation (sum of independent Poissons)

Let `X = (X1, X2)` with `X1 ~ Pois(2)`, `X2 ~ Pois(3)` independent, and define `Y = X1 + X2`.
This is implemented as a linear map with `A = [1 1]`.

```{r}
set.seed(1)

cg_vec <- PoissonCGF
A <- matrix(c(1, 1), nrow = 1)

cg_sum <- linearlyMappedCGF(cg_vec, A)

lambda <- c(2, 3)
Y <- cg_sum$rsim(iidReps = 10000, parameter_vector = lambda)

mean(Y)         # should be ~ 5
var(Y)          # should be ~ 5
```

## Example 3: Parameter mapping with adaptors (avoiding “wrong dimension” mistakes)

A common pitfall is passing a combined parameter vector to a component CGF that expects only one number.
For example, if `count_cgf` is `PoissonCGF` and you pass `theta = c(lambdaN, lambdaX)`,
then `PoissonCGF` interprets this as a **2-dimensional** Poisson parameter, not a scalar.

The fix is to *adapt* each sub-CGF so it receives only the piece of `theta` it needs.

```{r}
set.seed(1)

# theta = c(lambdaN, lambdaX)
theta <- c(lambdaN = 2, lambdaX = 3)

count_cgf   <- PoissonModelCGF(lambda = adaptor(indices = 1))
summand_cgf <- PoissonModelCGF(lambda = adaptor(indices = 2))

rss <- randomlyStoppedSumCGF(count_cgf, summand_cgf, block_size = 1)

Y <- rss$rsim(iidReps = 10000, parameter_vector = theta)
mean(Y)  # for compound Poisson, E[Y] = E[N] * E[X] = lambdaN * lambdaX = 6
```

## Example 4: RSS with vector summands (Poisson thinning check)

Let `N ~ Poisson(lambda)` and `X_i ~ Multinomial(1, p)` a one-hot vector (categorical draw).
Then `Y = sum_{i=1}^N X_i` should have independent Poisson components with means `lambda * p`.

```{r}
set.seed(1)

d <- 3
theta <- c(lambda = 10, p = c(0.2, 0.3, 0.5))

count_cgf <- PoissonModelCGF(lambda = adaptor(indices = 1))
summand_cgf <- MultinomialModelCGF(
  n        = adaptor(fixed_param = 1),
  prob_vec = adaptor(indices = 2:(d + 1))
)

rss_vec <- randomlyStoppedSumCGF(count_cgf, summand_cgf, block_size = d)

Y <- rss_vec$rsim(iidReps = 20000, parameter_vector = theta, drop = FALSE)

rowMeans(Y)  # should be ~ lambda * p = (2, 3, 5)
```

## Example 5: MultinomialCGF simulation (odds vs probabilities)

`MultinomialCGF` expects `parameter_vector = c(N, x_1, ..., x_d)`.
The `x` vector is treated as odds (or as probabilities if it already sums to 1).

```{r}
set.seed(1)

N <- 10
odds <- c(2, 3, 5)
param_odds <- c(N, odds)

X <- MultinomialCGF$rsim(iidReps = 6, parameter_vector = param_odds, drop = FALSE)
X
colSums(X)  # should all equal N

# Same distribution if you pass probabilities instead of odds
p <- odds / sum(odds)
param_p <- c(N, p)

Xp <- MultinomialCGF$rsim(iidReps = 6, parameter_vector = param_p, drop = FALSE)
colSums(Xp)
```

## Example 6: MultinomialModelCGF (theta-mapped multinomial)

```{r}
set.seed(1)

d <- 3
theta <- c(N = 12, p = c(0.1, 0.2, 0.7))

cgf <- MultinomialModelCGF(
  n        = function(th) th[1],
  prob_vec = function(th) th[2:(d + 1)],
  iidReps  = "any"
)

X <- cgf$rsim(iidReps = 5, parameter_vector = theta, drop = FALSE)
X
colSums(X)
```

## Example 7: SubunitaryMultinomialCGF

`SubunitaryMultinomialCGF` uses parameters `c(N, pi_1,...,pi_d)` where `sum(pi) <= 1`
and conceptually corresponds to a (d+1)-category multinomial **conditioned** on the last category being zero.

Simulation returns the `d` active-category counts and still sums to `N`.

```{r}
set.seed(1)

N <- 10
pi <- c(0.2, 0.3, 0.1)  # sum < 1 (subunitary)
param <- c(N, pi)

W <- SubunitaryMultinomialCGF$rsim(iidReps = 6, parameter_vector = param, drop = FALSE)
W
colSums(W)
```

## Example 8: sumOfiidCGF (sum of i.i.d. Poissons)

```{r}
set.seed(1)

base <- PoissonModelCGF(lambda = adaptor(indices = 1))
sum10 <- sumOfiidCGF(base, n = 10, block_size = 1)

theta <- c(lambda = 2)
Y <- sum10$rsim(iidReps = 20000, parameter_vector = theta)

mean(Y)  # should be ~ 20
var(Y)   # should be ~ 20
```



## Example: sumOfIndependentCGF (sum of independent, not-necessarily-identical terms)

```{r}
set.seed(1)

theta <- c(lambda1 = 2, lambda2 = 3)

cg1 <- PoissonModelCGF(lambda = adaptor(indices = 1))
cg2 <- PoissonModelCGF(lambda = adaptor(indices = 2))

cg_sum <- sumOfIndependentCGF(list(cg1, cg2), block_size = 1)

Y <- cg_sum$rsim(iidReps = 20000, parameter_vector = theta)

mean(Y)  # should be ~ 5
var(Y)   # should be ~ 5
```



## Other issues

### 1) RSS: `count_cgf` must be scalar

In RSS models, `count_cgf` represents a *scalar* count `N`, so simulation requires a scalar draw per replicate.
If your count CGF is vectorized (like `PoissonCGF`) and you accidentally pass a vector parameter,
you’ll get a “count must be scalar” error.

The fix is to use parameter adaptors (as shown above) so the count CGF receives only the scalar parameter it needs.

### 2) Esscher / exponential tilt wrappers typically do not implement `rsim()` (and what you can do instead)

`EsscherTiltCGF()` defines the **Esscher-tilted** CGF
\[
K_h(t) = K(t+h) - K(h),
\]
which corresponds to a change-of-measure
\[
\mathrm{d}\mathbb P_h(x) \;\propto\; e^{h^\top x}\,\mathrm{d}\mathbb P(x).
\]

This is not a deterministic transformation of draws from the base distribution.
In general, *even though the CGF determines the law (when it exists in a neighborhood of 0)*,
there is no universal, closed-form exact sampler that can be constructed from CGF evaluations alone.

Therefore, `EsscherTiltCGF()` is an 'analysis-only' (for saddlepoint / likelihood work) and
does not provide a generic exact `rsim()` implementation in this package. 

However, if you can simulate from the base model (i.e. `base_cgf$has_simulate() == TRUE`),
then you can still simulate **exactly/approximately** from the tilted distribution *outside the package*
using one of the following generic strategies:

- (A) Importance sampling + resampling (IS-R)  
  Draw many proposals \(X^{(m)} \sim \mathbb P\), weight them by \(w_m \propto e^{h^\top X^{(m)}}\),
  and resample.
- (B) Independence Metropolis–Hastings (IMH)  
  Use proposal \(q=\mathbb P\) (simulate from the base), and accept with probability
  \[
  \alpha(x\to x') = \min\{1,\exp(h^\top(x' - x))\}.
  \]
  This is attractive here because the unknown base density cancels in the MH ratio.

Both methods depend only on:
1) being able to draw from the base model via `base_cgf$rsim()`, and  
2) being able to compute \(h^\top x\) for simulated draws.

> **Note**  
> These are approximate samplers:
> IS-R has Monte Carlo error and can suffer weight degeneracy (low ESS), and
> IMH is asymptotically exact but has MCMC error and autocorrelation (needs diagnostics).

#### Helper functions (copy/paste)

```{r}
# Convert a draw to a d x B matrix (columns = draws)
.as_draw_matrix <- function(X) {
  if (is.null(dim(X))) {
    matrix(as.numeric(X), nrow = 1L)
  } else {
    X
  }
}

# Broadcast a scalar tilt to length d if needed (otherwise require exact match)
.broadcast_tilt <- function(tilt, d) {
  if (!is.numeric(tilt) || length(tilt) < 1L) stop("'tilt' must be a numeric vector.")
  if (any(!is.finite(tilt))) stop("'tilt' must be finite.")
  if (length(tilt) == 1L && d > 1L) return(rep(tilt, d))
  if (length(tilt) != d) stop("tilt length mismatch: expected length ", d, ", got ", length(tilt), ".")
  tilt
}

# Importance sampling + resampling
esscher_rsim_is_resample <- function(base_cgf,
                                     tilt,
                                     iidReps,
                                     parameter_vector,
                                     proposal_iidReps = max(2000L, 20L * iidReps),
                                     drop = TRUE,
                                     return_details = FALSE,
                                     ...) {
  if (!inherits(base_cgf, "CGF")) stop("'base_cgf' must be a CGF.")
  if (!isTRUE(base_cgf$has_simulate())) stop("Base CGF does not support simulation (has_simulate() is FALSE).")
  if (!is.numeric(iidReps) || length(iidReps) != 1L || iidReps < 1L) stop("'iidReps' must be a positive integer-ish.")
  iidReps <- as.integer(iidReps)
  proposal_iidReps <- as.integer(proposal_iidReps)

  Xprop <- .as_draw_matrix(base_cgf$rsim(iidReps = proposal_iidReps,
                                        parameter_vector = parameter_vector,
                                        drop = FALSE, ...))
  d <- nrow(Xprop)
  h <- .broadcast_tilt(tilt, d)

  # log-weights: lw_m = h^T x_m
  lw <- as.numeric(crossprod(h, Xprop))
  mx <- max(lw)
  if (!is.finite(mx)) stop("Non-finite tilt weights; try a smaller |tilt| or check your model.")

  w <- exp(lw - mx)
  sw <- sum(w)
  if (!is.finite(sw) || sw <= 0) stop("Degenerate importance weights; try more proposals or a smaller |tilt|.")
  w <- w / sw

  # Effective sample size (ESS) for diagnostics
  ess <- 1 / sum(w^2)

  idx <- sample.int(ncol(Xprop), size = iidReps, replace = TRUE, prob = w)
  out <- Xprop[, idx, drop = FALSE]

  if (drop && d == 1L) out <- as.numeric(out)

  if (isTRUE(return_details)) {
    return(list(draws = out, ess = ess, weights = w, logw = lw))
  }
  out
}

# Independence Metropolis–Hastings with proposal q = base distribution
esscher_rsim_imh <- function(base_cgf,
                             tilt,
                             iidReps,
                             parameter_vector,
                             burnin = 1000L,
                             thin = 1L,
                             init = NULL,
                             drop = TRUE,
                             return_details = FALSE,
                             ...) {
  if (!inherits(base_cgf, "CGF")) stop("'base_cgf' must be a CGF.")
  if (!isTRUE(base_cgf$has_simulate())) stop("Base CGF does not support simulation (has_simulate() is FALSE).")
  if (!is.numeric(iidReps) || length(iidReps) != 1L || iidReps < 1L) stop("'iidReps' must be a positive integer-ish.")
  iidReps <- as.integer(iidReps)
  burnin <- as.integer(burnin)
  thin <- as.integer(thin)
  if (burnin < 0L) stop("'burnin' must be >= 0.")
  if (thin < 1L) stop("'thin' must be >= 1.")

  # Infer dimension d from one base draw
  x1 <- .as_draw_matrix(base_cgf$rsim(iidReps = 1L, parameter_vector = parameter_vector, drop = FALSE, ...))
  d <- nrow(x1)
  h <- .broadcast_tilt(tilt, d)

  # Initialize chain
  if (is.null(init)) {
    x_curr <- as.numeric(x1)
  } else {
    init <- as.numeric(init)
    if (length(init) != d) stop("'init' must have length d = ", d, ".")
    x_curr <- init
  }

  n_iter <- burnin + thin * iidReps
  out <- matrix(NA_real_, nrow = d, ncol = iidReps)

  n_acc <- 0L
  k <- 0L

  for (iter in seq_len(n_iter)) {
    x_prop <- as.numeric(.as_draw_matrix(
      base_cgf$rsim(iidReps = 1L, parameter_vector = parameter_vector, drop = FALSE, ...)
    ))

    # log acceptance ratio: h^T(x_prop - x_curr)
    log_alpha <- sum(h * (x_prop - x_curr))

    # accept if log(U) < min(0, log_alpha)
    if (log(runif(1)) < min(0, log_alpha)) {
      x_curr <- x_prop
      n_acc <- n_acc + 1L
    }

    if (iter > burnin && ((iter - burnin) %% thin == 0L)) {
      k <- k + 1L
      out[, k] <- x_curr
    }
  }

  acc_rate <- n_acc / n_iter
  if (drop && d == 1L) out <- as.numeric(out)

  if (isTRUE(return_details)) {
    return(list(draws = out, accept_rate = acc_rate))
  }
  out
}
```

#### Worked example: approximate tilted simulation + saddlepoint MLE

We use a *simple* base model (Poisson) with a known closed-form tilt.
The exact tilted law is:
if \(X \sim \mathrm{Pois}(\lambda)\), then under tilt \(\tau\),
\(X^{(\tau)} \sim \mathrm{Pois}(\lambda e^\tau)\).

Even though we could sample exactly here, the *same IS-R / IMH code* applies
to more complex composite CGFs whenever `base_cgf$rsim()` exists.

```{r}
set.seed(1)

# Base model: Poisson with theta = lambda
theta_true <- c(lambda = 10)
tau <- 0.4

base <- PoissonModelCGF(lambda = adaptor(indices = 1))
tilted_cgf <- EsscherTiltCGF(base, tilt = tau)

# Theory check: mean under the tilt is K'_base(tau) = lambda * exp(tau)
mean_theory <- as.numeric(base$K1(tau, theta_true))

# --- (A) Importance-resampling sampler ---
sim_is <- esscher_rsim_is_resample(
  base_cgf = base,
  tilt = tau,
  iidReps = 2000,
  parameter_vector = theta_true,
  proposal_iidReps = 20000,
  return_details = TRUE
)

mean(sim_is$draws)
mean_theory
sim_is$ess  # diagnostic; higher is better (<= proposal_iidReps)

# --- (B) Independence MH sampler ---
sim_mh <- esscher_rsim_imh(
  base_cgf = base,
  tilt = tau,
  iidReps = 2000,
  parameter_vector = theta_true,
  burnin = 2000,
  thin = 2,
  return_details = TRUE
)

mean(sim_mh$draws)
mean_theory
sim_mh$accept_rate
```

Now we fit \(\lambda\) by saddlepoint MLE under the tilted model,
using the simulated tilted sample as data.

```{r}
# Using the IS-R draws as "observed data" under the tilted model
y_obs <- sim_is$draws

fit <- find.saddlepoint.MLE(
  observed.data  = y_obs,
  cgf            = tilted_cgf,
  starting.theta = c(lambda = 1),
  lb.theta       = c(lambda = 1e-8),
  ub.theta       = c(lambda = 100),
  method         = "two_step",
  std.error      = FALSE
)

fit$MLEs.theta
theta_true
```

> **Notes**
>
> - For high-dimensional \(X\), IS-R often degenerates unless the tilt is small or the proposal sample size is very large (check ESS).
> - IMH should be more robust than IS-R when weights explode, but you must check mixing / autocorrelation (acceptance rate is only one diagnostic).
> - If your base family has a closed-form tilted law (Normal / Poisson / Gamma / ...), prefer that exact sampler. The generic IS-R / IMH methods are fallbacks.


---

## Adding a new `rsim()` implementation

For any new base distribution, define a simulator and pass it to `createCGF(rsim = ...)`.

**Contract expected by `CGF$rsim`:**

- simulator returns either:
  - a numeric vector of length `d * iidReps`, or
  - a numeric `d x iidReps` matrix.
- each column corresponds to one simulated draw of the underlying random vector.

Then wrappers can often be implemented in a small, compositional way:
e.g. `linearlyMappedCGF` simulates `X` and returns `A %*% X`.

